Modularly Composing Typed Language Fragments

### 2014-11-02 att-pldi15/att-pldi15-supplement.tex
Finished draft of PLDI15 except some citations

\title{Modularly Composing Typed Language Fragments}

\authorinfo{}{}{}
%\authorinfo{Cyrus Omar \and Jonathan Aldrich}
 %         {Carnegie Mellon University}
  %         {\{comar, aldrich\}@cs.cmu.edu}

\maketitle
\begin{abstract}
Researchers often describe typed programming languages as fragments or simple calculi, leaving to language designers the task of composing these to form complete languages.  %comparable to those available for library-based embeddings packaged using a modern module system. 
This is not a systematic process: metatheoretic results must be established anew for each composition, guided only notionally by metatheorems derived for simpler systems.
As the design space grows, mechanisms that provide stronger modular reasoning principles than this are needed.%, so that important metatheoretic results need only be established locally and, ultimately, so that language extensions can be treated like libraries. 
%,  Sometimes this fails because certain language features do not ``play well'' together, but precisely characterizing which is elusive given these practices.

%Recent work has started to address this problem, e.g. by showing how to modularly reason about desugarings atop a fixed type system. Our focus here is on safely composing type system fragments. We organize fragments around type constructors, as is usual practice when describing type systems, and sidestep the difficulties of composing abstract syntax (i.e. the expression problem)  by delegating control over a small, uniform abstract syntax in a type-directed manner to logic associated with a type constructor.

In this paper, we begin from first principles with a core calculus, @$\lambda$, specified like many full-scale languages: as a bidirectionally typed translation semantics. Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show a variant of record types and constrained string types) are defined by extending a \emph{tycon context}. Each tycon defines the semantics of associated term-level operators (e.g. record projection) using functions written in a static language where types and translations are values. The semantics provide  strong metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that all \emph{tycon  invariants} will be conserved under extension. Mechanized proofs are not needed: problems are caught during typechecking by lifting typed compilation techniques into the semantics and enforcing abstraction barriers around tycons using type abstraction, the same principle that underlies ML-style modules.
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%extensible languages; module systems; type abstraction; typed compilation; type-level computation
\section{Introduction}\label{intro}
Typed programming languages are most often described as being composed of \emph{fragments}, each contributing to the language's concrete syntax, abstract syntax, static semantics and dynamic semantics. 
In his textbook, Harper organizes fragments around type constructors, describing each in a different chapter \cite{pfpl}. Languages are then identified by a set of type constructors, e.g. $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language that includes partial function types, polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed below).
Another related approach is to describe fragments with a simple calculus having a ``catch-all'' constant and base type to stand notionally for all other terms and types that may also be included in some future complete language.

In contrast, the usual metatheoretic reasoning techniques for programming languages  (e.g., rule induction) operate on complete language specifications. Each {combination} of fragments must formally be treated as its own monolithic language for which  metatheorems must be established anew, guided only informally by those derived for the smaller systems from which the language is notionally composed.

%In both cases, it is left as an extrinsic concern to ensure that the metatheory developed separately is actually conserved when fragments defined in these ways are composed to form a complete language.

This is not an everyday problem for programmers only because fragments like those mentioned above are ``general purpose'': they make it possible to \emph{isomorphically embed}  many other fragments as ``libraries''. For example, list types need not be built in because they are isomorphic to the type $\iforall{\alpha}{\imu{t}{\iunit + (\alpha \times t))}}$ (datatypes in ML combine these into a single declaration construct). %Languages providing datatypes al\'a ML  are perhaps most directly oriented around  embeddings into this core. % Leaving the core language simple makes it easier to establish its metatheory and verify that tools, like compilers, are implemented correctly.% They also have encouraging connections to logic. %For example, we do not need to define the type constructor $\fvar{list}$ (indexed by a type) as a fragment (though it is possible to do so) because there is a user-defined datatype constructor, $\texttt{list}$, parameterized by a type, that is isomorphic. If we added the $\fvar{list}$ fragment to the language, it would be entirely redundant semantically: every well-typed term of a type constructed by $\fvar{list}$ corresponds to a well-typed term and a polymorphic recursive sum type such th%Although strings and numbers can be embedded as recursive sum types, it is recognized that this is impractical, so these are also usually included as primitives.
%Establishing an isomorphic embedding of a desirable fragment in terms of general-purpose fragments is not always possible, nor are such embeddings always practical.

Universality properties (i.e. ``Turing-completeness'') can be invoked to guarantee that an embedding that preserves a desirable fragment's dynamic semantics is possible, but an isomorphic embedding must also preserve the static semantics and, if defined, performance bounds specified using a cost semantics. This is not always possible. 
Embeddings are also sometimes too \emph{complex}, as measured by the cost of the extralinguistic computations that are needed to map in and out of the embedding and, if these  must be performed mentally by programmers, considering various human factors. %We will discuss specific examples below.
%When an embedding is too complex, abstraction providers have a few options, discussed in Sec. \ref{prior-work}. When an embedding is not possible, however, or when these options are  insufficient, 
%providers are often compelled to introduce these fragments by extending an existing language, thereby forming a new \emph{dialect}. To save effort, they may do so by forking existing artifacts or leverage tools like compiler generators or language frameworks, also reviewed in Sec. \ref{prior-work}. %Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped. 
Each time a language is extended with a new fragment for one of these reasons, a new \emph{dialect} is born. Within the ML lineage, for example, dialects abound:
%Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
%\begin{quote}
%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
%\end{quote}
%

\begin{enumerate}
\vspace{-5px}
%\compresslist
\item 
\textbf{General Purpose Fragments:} 
A number of variations on product types, for example, have been introduced in dialects: 
$n$-ary tuples, 
labeled tuples, 
records (identified up to reordering), 
records with width and depth subtyping \cite{Cardelli:1984:SMI:1096.1098}, 
records with update operators\footnote{The Haskell wiki notes that ''No, extensible records are not implemented in GHC. The problem is that the record design space is large, and seems to lack local optima. [...] As a result, nothing much happens.'' \cite{GHCFAQ}} \cite{ocaml-manual}, 
records with mutable fields \cite{ocaml-manual}, 
%field delegation \cite{atlang-gpce14} \todo{gpce submission} 
and 
records with ``methods'' (i.e. pure objects \cite{TSLs}). 
 Sum-like types are also exposed in various ways: 
finite datatypes, 
open datatypes \cite{conf/ppdp/LohH06}, 
hierarchically open datatypes \cite{journals/toplas/MillsteinBC04}, 
polymorphic variants \cite{ocaml-manual} and 
ML-style exception types. Combinations of these manifest themselves as class-based object systems \cite{ocaml-manual}.

\item
\textbf{Specialized Fragments:} Fragments that track specialized static invariants to provide stronger correctness or security guarantees, manage unwieldy lower-level abstractions and run-time systems or control cost are also frequently introduced in dialects, e.g. for data parallelism  \cite{chakravarty2007data}, distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, reactive programming \cite{mandel2005reactiveml}, authenticated data structures \cite{Miller:2014:ADS:2535838.2535851}, databases \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09} and regular string sanitation \cite{sanitation-psp14}. %\verb|sprintf| in the OCaml dialect statically distinguishes format strings from strings.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.

\item
\textbf{Foreign Fragments:} A safe and natural foreign function interface (FFI) can be valuable (particularly given this proliferation of dialects). This requires enforcing the type system of the foreign language in the calling language. %Using  a FFI that does not do this can lead to safety issues, even when both languages are separately known to be safe. 
%Safe FFIs generally require direct extensions to the language. 
For example, MLj builds in a safe FFI to Java \cite{Benton:1999:IWW:317636.317791}.
\end{enumerate}
\vspace{-5px}

This \emph{dialect-oriented} state of affairs is unsatisfying. %Language designers are burdened with needing to understand the complete metatheory of every fragment they add to their language to make sure it is not incompatible with existing fragments, so languages change rarely. This decreases the impact of many potentially useful innovations designed by fragment providers (e.g. many academic researchers) by making them unavailable to programmers. 
While programmers can choose from dialects supporting, e.g., a principled approach to distributed programming, or one that builds in support for statically reasoning about units of measure, one that supports both fragments may not be available. Using different dialects separately for different components of a program is untenable: components written in different dialects cannot always interface safely (i.e. a safe FFI, item 3 above, is needed).

These problems do not merit concern when a fragment is expressed as an isomorphic embedding (i.e. as a library) because modern \emph{module systems} enforce abstraction barriers that ensure that the isomorphism need only be established in the ``closed world'' of the module. For example, a module defining sets in ML can hold the representation of sets abstract, ensuring that any invariants maintained by the functions in the module (e.g. uniqueness, if using a list representation) will hold no matter which other modules are separately in use by a client.\todo{citation?} %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec. \ref{prior-work} by summarizing them.

When library-based embeddings are not possible, as in the examples above, mechanisms are needed that make it possible to define and  reason in a similarly modular manner about  direct extensions to the semantics of a language. Such a mechanism could ultimately be integrated directly into the language, blurring the distinction between fragments and libraries and decreasing the need for new dialects.% Importing fragments that introduce new semantics   would be as easy and reasonable as importing a new module is in ML today. %In the limit, the community could rely on modularly mechanized metatheory and compiler correctness results.% rather than requiring heroic efforts from individual research groups that consider an entire language at once. %Recent work has shown progress on modularly introducing new concrete syntax, reviewed in Sec. \ref{desugaring}. Our focus is on the problem of introducing new semantics.

\paragraph{Contributions} In this paper, we take foundational steps towards this goal by  constructing a simple but surprisingly powerful core calculus, @$\lambda$ (the ``actively typed'' lambda calculus). %Despite its minimality, it can host a variety of practical semantic extensions like those described above, while maintaining strong metatheoretic guarantees and, crucially, providing modular reasoning properties. %a calculus introduced briefly in recent work on \emph{active type constructors}  \cite{atlang-gpce14}\todo{TR? Arxiv?}, reviewed in Sec. \ref{overview}. 
Its semantics are structured like those of many modern languages, consisting of an \emph{external language} (EL) governed by a {typed translation semantics} targeting a much simpler \emph{internal language} (IL). 
Rather than building in a monolithic set of external type constructors, however, the  semantics are indexed by a \emph{tycon context}. Each tycon defines the semantics of its operators via functions written in a \emph{static language} (SL) where types and translations are values. %, which provides a form of type-level computation. %The authors demonstrate the expressive power of this technique primarily with an implementation and a number of examples of fragments as libraries, and their core calculus .% (unlike systems that treat the type system as a ``bag of rules'', where non-determinism can arise).

We will begin by giving an overview of the organization  and main judgements of @$\lambda$ in Sec. \ref{atlam}, then discuss how types are constructed and introduce our two main examples, one defining labeled product types with a functional update operator, and the other regular string types, based on a recent core calculus style specification \cite{sanitation-psp14}, in Sec. \ref{types}. We describe how tycons control the semantics of their associated term-level operator constructors (opcons) in Sec. \ref{external-terms}.  We next give the key metatheoretic properties of the calculus in Sec. \ref{metatheory}, including \emph{type safety} and a key modularity result, which we call  \emph{conservativity}: any invariants that can be established about all values of a type under \emph{some} tycon context (i.e. in some  ``closed world'') are conserved in any further extended tycon context (i.e. in the ``open world''). Interestingly, type system providers need not necessarily provide mechanized proofs to maintain these guarantees. Instead, the approach we take relies on type abstraction in the internal language. As a result, we are able to borrow the same results that underly modular reasoning in simply-typed languages like ML to reason modularly about typed language fragments. We describe  related work in Sec. \ref{prior-work} and conclude with a brief discussion of expressiveness in Sec. \ref{discussion}.

\section{Overview of @$\lambda$}\label{atlam}\label{overview}
\paragraph{External Language}
Programmers interface with @$\lambda$ by writing \emph{external terms}, $e$. The abstract syntax of external terms is shown in Figure \ref{syntax-EL} (we will give various concrete desugarings in Sec. \ref{external-terms}). %We will describe useful syntactic desugarings atop this syntax as we go on and review recent techniques that permit modularly introducing  desugarings like these in Sec. \ref{desugaring}. Our main focus here is on semantic extensions. 
The static and dynamic semantics are given simultaneously as a \emph{bidirectionally typed translation semantics}, i.e. the key judgements take the form:  \[\esynX{e}{\st\moutput}{\iota\moutput} ~~~~~\text{and}~~~~~ \eanaX{e}{\st}{\iota\moutput}\]
\noindent

These are pronounced ``$e$ (synthesizes / analyzes against) type $\sigma$ and has  translation $\iota$ under typing context $\Upsilon$ and tycon context $\Phi$''. Note that our specifications in this paper are intended to be algorithmic: we indicate ``outputs'' when introducing judgement forms by \emph{mode annotations}, $\moutput$; these are not part of the judgement's syntax. In particular, note that the type is an ``output'' only in the synthetic judgement.

This basic separation of the EL and IL is commonly used for full-scale language specifications, e.g. the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic}. The internal language is purposely kept small, e.g. defining only simple products, to simplify metatheoretic reasoning and compilation. The EL then specifies various useful higher-level constructs, e.g. record types, by translation to the IL. In @$\lambda$, the EL builds in only function types. All other external constructs are defined in the \emph{tycon context}, described in the Sec. \ref{types}.

We choose bidirectional typechecking, also sometimes called \emph{local type inference} \cite{Pierce:2000:LTI:345099.345100}, for two main reasons. The first is once again to justify the practicality of our approach: local type inference is increasingly being used in modern languages (e.g. Scala\todo{citation}) because it eliminates the need for type annotations in many situations while remaining decidable in more situations than whole-function type inference and providing what are widely perceived to be higher quality error messages\todo{citation from Dunfield}. Secondly, it will give us a clean way to reuse the generalized introductory form, $\eintro{\st}{\es}$, and its associated desugarings, at many types. For example, regular string types will be able to use standard string literal syntax.

Unlike the Harper-Stone semantics, where external and internal terms were governed by a common type system, in @$\lambda$ each external type, $\st$, maps onto an internal type, $\tau$, called the \emph{type translation} of $\st$. This mapping is specified by the  type translation judgement, $\vdash_\Phi \st \leadsto \tau$, which will be described in Sec. \ref{sec:type-translations}. 
For this reason, this specification style may also be compared to specifications for the first stage of a type-directed compiler, e.g. the TIL compiler for Standard ML \cite{tarditi+:til-OLD}, here lifted ``one level up'' into the semantics of the language itself. As we will see, type safety follows from a property analagous to a correctness condition that arises in typed compilers. Modular reasoning will be based on holding the type translation of $\st$ abstract ``outside'' the tycon.

%We will return to how we control the semantics of the single introductory form in the abstract syntax, $\eintro{\st}{\es}$, on the basis of the type it is being analyzed against, sidestepping issues related to the \emph{expression problem} as a result \todo{cite other paper}\cite{wadler1998expression}. The form $\eother{\iota}$ is a technical device used to quantify over all terms that may arise in a ``future'' tycon context, serving an analagous role to the  ``catchall'' base constants commonly found in minimal calculi, and we will return to its semantics later as well.%We will see how we leverage this bidirectionality when we discuss literal forms below. % The judgement is only well-defined for \emph{valid} tycon contexts, written judgementally as $\vdash \Phi$ and \emph{valid} typing contexts, $\vdash_\Phi \Upsilon$, which we also specify below.
%  \begin{figure*}[t!]
% \small
% \hspace{-5px}\input{example-table}
% %\\~\\
% %$\begin{array}{rcl}
% %e\cdot\conclbl{lbl}(\conclbl{lbl}_1=e_1, \cdots, \conclbl{lbl}_n=e_n)  :=  \etarg{(\conclbl{lbl}; \svar{list}n[\klbl]~\conclbl{lbl}_1~\cdots~\conclbl{lbl}_n)}{e}{e_1; \cdots; e_n}
% %\end{array}$
% \caption{An example external term, $e_{ex}$, in concrete syntax (left), desugared to abstract syntax (right), with static terms shown in green in examples (only). The notation ${\small \scolor{\desugar{...}}}$ stands for an embedding of the indicated ${\small \scolor{\conclbl{label}}}$, ${\small \scolor{\concrx{regular expression}}}$, ${\small \scolor{\concstr{string}}}$ or numeral into the SL, with derived kinds  $\klbl$, $\krx$,  $\kstr$ and $\knat$, not shown. The signatures of helper functions, e.g. ${\small \scolor{\svar{nil}}}$ and ${\small \scolor{\svar{list}n}}$, are shown in Figure \ref{helper-sigs}.}
% \label{example}
% \end{figure*}
\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-IL}
\caption{Syntax of {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, our internal language (IL). Metavariable $x$ ranges over term variables and $\alpha$ and $t$ both range over type variables.}
\label{syntax-IL}
\vspace{-10px}
\end{figure}

\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-EL}
\caption{Syntax of the external language (EL).}\label{syntax-EL}
\end{figure}

\paragraph{Internal Language} 
%Internal terms, $\iota$, together with internal types, $\tau$, form the \emph{typed internal language}. 
@$\lambda$ relies on a typed internal language supporting type abstraction (i.e. universal quantification over types). We use {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, the syntax for which is shown in Figure \ref{syntax-IL}, as representative of any intermediate language for a typed functional language. %Note that the IL has a fixed semantics; changes to the IL induce different dialects of @$\lambda$.

We assume the statics of the IL are specified in the standard way by judgements for  type formation {$\iDelta \vdash \tau$}, typing context formation { $\iDelta \vdash \iGamma$} and type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}. 
%The typing and type formation contexts obey standard structural properties (i.e. weakening, exchange and contraction). 
In examples, we will omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences (e.g. writing $\Gamma_\text{test} := x : \tau$).

The internal dynamics are specified as a structural operational semantics with a stepping judgement {\small $\iota \mapsto \iota\moutput$} and a value judgement {$\iota~\mathtt{val}$}. The multi-step judgement $\iota \mapsto^{*} \iota\moutput$ is the reflexive, transitive closure of the stepping judgement and the evaluation judgement $\iota \Downarrow \iota'$ is defined iff $\iota \mapsto^{*} \iota'$ and $\iota'~\mathtt{val}$. Both the static and dynamic semantics of the IL can be found in any standard textbook covering typed lambda calculi (we directly follow \cite{pfpl}), so we assume familiarity and key metatheoretic properties.

% We also here define a syntax for simultaneous substitutions, of terms for variables, $\gamma$, and of types for polymorphic type variables, $\delta$, and assume standard judgements ensuring that these are valid with respect to a valid context, $\Delta \vdash \gamma : \Gamma$ and $\vdash \delta : \Delta$. We apply these substitutions to terms, types and typing contexts using the syntax $[\gamma]\iota$, $[\delta]\iota$, $[\delta]\tau$ and $[\delta]\Gamma$ (in some prior work, application of a substitution like this is written using a hatted form, e.g. $\hat\gamma(\iota)$; we intend the same semantics but use a notation more consistent with standard substitution, e.g. $[\iota/x]\iota'$). We will omit  leading $\emptyset$ and $\cdot$ in examples; in specifications, the former is used for  metatheoretic finite mappings, the latter for metatheoretic ordered lists.

\paragraph{Static Language}
\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-SL}
\caption{Syntax of the static language (SL). Metavariable $\sx$ ranges over static term variables, $\kalpha$ and $\kvar$ over kind variables and $m$ and $n$ over natural numbers.}\vspace{-5px}
\label{syntax-SL}
\end{figure}
The workhorse of @$\lambda$ is the \emph{static language}, which itself forms a typed lambda calculus where 
%External terms are classified by (external) \emph{types}. 
\emph{kinds}, $\kappa$, classify \emph{static terms}, $\sigma$.  The syntax of the SL is given in Figure \ref{syntax-SL}. The portion of the SL covered by the first row of kinds and first four rows of static terms in the syntax forms an entirely standard  functional programming language consisting of total functions, universal quantification over kinds, inductive kinds (constrained by the standard positivity condition to prevent non-termination), and products and sums. The reader can consider these as forming a total subset of ML or a simply-typed subset of Coq. The semantics also directly follow \cite{pfpl}, so we  omit the details of this core here.  Only three new  kinds will be needed: $\kty$ (Sec. \ref{types}), $\kity$ (Sec. \ref{sec:type-translations}) and $\kitm$ (Sec. \ref{sec:introop}).

The kinding judgement takes the form $\sofkX{\st}{\kappa\moutput}$, where $\kDelta$ and $\kGamma$ are analagous to $\iDelta$ and $\iGamma$ and analagous kind and kinding context formation judgements $\kDelta \vdash \kappa$ and $\kDelta \vdash \kGamma$ are  defined. The natural number $n$ is used as a technical device in our semantics to ensure that the forms shown as being indexed by $n$ in the syntax only arise in a controlled manner internally to prevent ``out of bounds'' issues, as we will discuss; they would have no corresponding concrete syntax so $n$ can be assumed $0$ in user-defined terms.

The dynamic semantics of static terms is defined as a structural operational semantics by a stepping judgement $\sstep{\st}{\argEnv}{\st\moutput}$, a value judgement $\sval{\st}{\argEnv}$ and an error judgement $\serr{\st}{\argEnv}$. Here, $\argEnv$ ranges over \emph{argument environments}, which we will  return to when considering opcons in Sec. \ref{external-terms}. The multi-step judgement $\smanystep{\st}{\argEnv}{\st\moutput}$ is the reflexive, transitive closure of the stepping judgement. The normalization judgement $\seval{\st}{\argEnv}{\st'}$ is defined iff $\smanystep{\st}{\argEnv}{\st'}$ and $\sval{\st'}{\argEnv}$. %${\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$, where $\memD$, $\memG$ and $\aCtx$ are also technical devices that will be described later; they too can be ignored from the perspective of user code. We write $\st \Downarrow \st'$ iff $\snorm{\st}{\emptyset}{\cdot}{{\rightharpoonup}; \emptyset; \emptyset}{\st'}{\emptyset}{\cdot}$. \emph{Static values} are normalized static terms. Normalization can also raise an error (to indicate a type error in an external term, or a problem in a tycon definition, as we will discuss), indicated by the judgement $\serrX{\st}$. We omit error propagation rules.%We will refer to the relevant rules as we proceed. 
\begin{figure}[t]
\small
$\begin{array}{lrcl}
\textbf{tycons} & c & ::= & \rightharpoonup ~|~ \tc ~|~ \keyw{other}[m]\\
\textbf{tycon contexts} & \Phi & ::= & \cdot ~|~ \Phi, \tcdef{\tc}{\psi}{\theta}\\
\textbf{tycon structures} & \theta & ::= & \tcstruct{\st}{\omega} \\
\textbf{tycon sigs} & \psi & ::= & \tcsig{\kappa}{\chi}\\
\textbf{opcon structures} & \omega & ::= & \tcstructn{\st} ~|~ \tcstructc{\omega}{op}{\st}\\
\textbf{opcon sigs}& \chi & ::= & \introsig{\kappa} ~|~ \opsigS{\theta}{op}{\kappa}\\
\end{array}$
\caption{Syntax of tycons. Metavariables $\tc$ and $\opname{op}$ range over user-defined tycon and opcon names, respectively, and $m$ ranges over natural numbers.}
\label{syntax-TC}
\end{figure}
\begin{figure*}\begin{mathpar}
\small
\inferrule[k-parr]{
    \sofkX{\st}{\kprod{\kty}{\kty}}
}{
    \sofkX{\sty{\rightharpoonup}{\st}}{\kty}
}

\inferrule[k-ty]{
    \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi\\
    \sofkX{\st}{\ktyidx}
}{
    \sofkX{\sty{\tc}{\st}}{\kty}
}

\inferrule[k-otherty]{
    \sofkX{\sttyidx}{\keyw{Nat} \times \kity}
}{
    \sofkX{\sty{\keyw{other}[m]}{\sttyidx}}{\kty}
}\vspace{-10px}
\end{mathpar}
\caption{Kinding rules for types, which take the form $\sty{c}{\sttyidx}$ where $c$ is a tycon and $\sttyidx$ is the type index.}
\label{fig:types}
\end{figure*}

\section{Types}\label{types}

External types, or simply \emph{types}, are static values of kind $\kty$. The introductory form for kind $\kty$ is $\sty{c}{\st}$, where $c$ is a \emph{tycon} and $\st$ is the \emph{type index}. The syntax  for tycons given in Figure \ref{syntax-TC} specifies that $c$ is either the built-in tycon governing partial functions, $\rightharpoonup$, a user-defined tycon name written in small caps, $\tc$, or an ``other'' tycon, $\keyw{other}[m]$ indexed by a natural number $m$ (to allow arbitrarily many such tycons). The kinding rules governing the form $\sty{c}{\st}$ are shown in Figure \ref{fig:types}. The dynamics are simple (see supplement): the index is eagerly normalized and errors propagate. We write $\istype{\st}{\Phi}$  iff $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\kty}$ and $\svalNA{\st}$.

The rule \rulename{k-parr} specifies that the type index of partial function types must be a pair of types. We thus say that $\rightharpoonup$ has \emph{index kind} $\kty \times \kty$. To recover a conventional syntax, we can introduce a desugaring from $\st_1 \rightharpoonup \st_2$ to $\sty{\rightharpoonup}{(\st_1, \st_2)}$. %Here, we will benefit by treating it uniformly.

All user-defined tycons must be defined in the \emph{tycon context}, $\Phi$, which is simply a list of tycon definitions. Each tycon defines a name, $\tc$, a \emph{tycon structure}, $\theta$, and a \emph{tycon signature}, $\psi$. We will return to the tycon structure below. Tycon signatures have the form $\tcsig{\ktyidx}{\chi}$, where $\ktyidx$ is the tycon's index kind and $\chi$ is the \emph{opcon signature}, which we   discuss in Sec. \ref{external-terms}. The first premise of \rulename{k-ty} extracts the index kind and the second checks the type index against it.

The rule (k-otherty) governs types constructed by an ``other'' tycon. These will serve only as technical devices to stand in for types other than those in a given tycon context in Sec. \ref{metatheory}. The index of such a type must pair a natural number with a type translation of kind $\kity$, discussed in Sec. \ref{sec:type-translations}.

\paragraph{Examples}
As our first example, consider the user-defined tycon $\tcvar{rstr}$. The index kind of $\tcvar{rstr}$ is $\krx$, which classifies static regular expression patterns and we assume is defined as an inductive sum kind in the usual way. Types constructed by $\tcvar{rstr}$ will classify \emph{regular strings}, which are statically known to be in the regular language specified by the type index  \cite{sanitation-psp14}. For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$ will classify non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ will classify conference names.  The type indices are  here written using standard concrete syntax for concision. Previous work has shown how to define type-specific (here, kind-specific) syntax like this composably in libraries \cite{TSLs}. We define a tycon context containing only the definition of $\tcvar{rstr}$, $\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\tcsig{\krx}{\chi_\text{rstr}}}{\theta_\text{rstr}}$.

Our second example is the tycon $\tcvar{lprod}$, which will define a variant of labeled product type (labeled products are like record types, but maintain a row  ordering; record types are also definable in a manner discussed in the supplement, but maintaining an ordering simplifies our discussion). We choose the index kind of $\tcvar{lprod}$ to be $\klist{\klbl \times \kty}$, where list kinds are defined as inductive sums in the usual way, and $\klbl$ classifies static representations of row labels. The tycon context containing only $\tcvar{lprod}$'s definition is $\small\Phi_\text{lprod} := \tcdef{\tcvar{lprod}}{\tcsig{\klist{\klbl \times \kty}}{\chi_\text{lprod}}}{\theta_\text{lprod}}$.

In the tycon context containing both tycon definitions, $\Phi_\text{rstr}\Phi_\text{lprod}$, we can define a labeled product type classifying conference papers, $\stx{paper} := \sty{\tcvar{lprod}}{\{\conclbl{title} : \stx{title}, \conclbl{conf} : \stx{conf}\}}$. Note that $\istype{\stx{paper}}{\Phi_\text{rstr}\Phi_\text{lprod}}$ and we again use kind-specific syntax, in this case for $\klbl$ and $\klist{\klbl \times \kty}$.

\subsection{Static Type Case Analysis}
A type $\st$ can be case analyzed using $\stycase{c}{\st}{\sx}{\st_1}{\st_2}$. If the value of $\st$ is constructed by $c$, its type index is bound to $\sx$ and the branch $\st_1$ is taken. For totality, a default branch, $\st_2$, must also be provided.  For example, the kinding rule for when $c$ is user-defined is below. 
\begin{mathpar}
\small
\inferrule[k-tycase]{
    \sofkX{\st}{\kty}\\
    \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi\\
    \sofk{\kDelta}{\kGamma, \sx :: \ktyidx}{\Phi}{\st_1}{\kappa}\\
    \sofkX{\st_2}{\kappa}
}{
    \sofkX{\stycase{\tc}{\st}{\sx}{\st_1}{\st_2}}{\kappa}
}
\end{mathpar}
% \begin{mathpar}\small
% \inferrule[s-tycase-fail-2]{ }{
%     \sstep{\stycase{c}{\sotherty{m}{\tau}}{x}{\st_1}{\st_2}}{\argEnv}{\st_2}
% }
% \end{mathpar}
The rule for $c{=}{\rightharpoonup}$ is analagous, but no rule for $c{=}\keyw{other}[m]$ is defined (these always take the default branch).

Put another way, types can be thought of as arising from a distinguished ``open datatype'' defined by the tycon context\todo{citation}. %The dynamics (see supplement) are again straightforward, consistent with this intuition.% We will see an example of its use in Sec. \ref{sec:targops}.
\begin{figure}\hfill \fbox{$\vdash \Phi$}\vspace{-25px}\begin{mathpar}
\small
\inferrule[tcc-ext]{
    \vdash \Phi\\
    \tc \notin \text{dom}(\Phi)\\
    \keq{\emptyset}{\ktyidx}\\
    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\\\
    \vdash_{\Phi,  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}
}{
    \vdash \Phi, \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}
}\vspace{-8px}
\end{mathpar}
\caption{Tycon context well-definedness. We omit the trivial case for when $\Phi=\cdot$ for concision.}
\label{fig:tycon-ctxs}
\vspace{-5px}
\end{figure}
\subsection{Tycon Context Well-Definedness}
The tycon context well-definedness judgement, $\vdash \Phi$, shown in Figure \ref{fig:tycon-ctxs}, requires that all tycon names are unique and performs additional checks,  described below.

\subsection{Type Equivalence}\label{sec:type-equivalence}
The first check simplifies the handling of type equivalence: type index kinds must be \emph{equality kinds}, i.e. those for which semantic equivalence implies syntactic equality at normal form. We define these by the judgement $\keq{\kDelta}{\kappa}$ (see supplement). 
Equality kinds are similar to equality types as found in Standard ML\todo{citation}. The main implication of this choice is that type indices cannot contain static functions.

\subsection{Type Translations}\label{sec:type-translations}
Each tycon in the tycon context {computes} translations for the types it constructs as a function of each type's index by specifying a \emph{translation schema} in the {tycon structure}, $\theta$. A tycon with index kind $\ktyidx$ defines a translation schema of kind $\ktyidx \rightarrow \kity$, checked by (tcc-ext).

The kind $\kity$ has a single introductory form, $\sqity{\qity}$, where $\qity$ is a \emph{translational internal type}. Each form in the syntax for internal types, $\tau$,  corresponds to a form in the syntax of translational internal types, $\qity$. For example, our translation schema for $\tcvar{rstr}$  simply chooses to ignore the type index and represent all regular strings internally as strings, of internal type abbreviated $\keyw{str}$. We abbreviate the corresponding translational internal type $\hat{\keyw{str}}$ and define the translation schema as $\small\theta_\text{rstr} := \tcstruct{\slam{\krx}{\svar{tyidx}}{\sqity{\hat{\keyw{str}}}}}{\omega_\text{rstr}}$.

>>>>

\title{Modularly Composing Typed Language Fragments}
\subtitle{Supplemental Material}
\authorinfo{}{}{}
%\authorinfo{Cyrus Omar \and Jonathan Aldrich}
 %         {Carnegie Mellon University}
  %         {\{comar, aldrich\}@cs.cmu.edu}

\maketitle

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%extensible languages; module systems; type abstraction; typed compilation; type-level computation

\section{Internal Language}
\todo{copy from one of the 312 HWs}

\subsection{Substitutions}

\subsection{Abstraction Theorem}
\todo{hm...}

\section{Tycon Contexts}
\subsection{Tycon Context Well-Definedness}

\subsection{Equality Kinds}
Need an equational theory for SL to state equality kind property, but not important for other metatheory.

\subsection{Full Examples}

\section{Static Language}
\subsection{Kind Formation}

\subsection{Kinding Context Formation}

\subsection{Kinding}

\subsection{Dynamic Semantics}

\subsection{Kind Safety}

\section{Types}
\subsection{Type Translations}

\subsection{Typing Context Translations}
